---
title: "CHL5222 Group 5 Final Project"
subtitle: "Feature Engineering, Primary Analysis, Secondary Analysis, Sensitivity Analysis, Missing Data Analysis, and Model Validation"
author: "Shirling Liu"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  out.width = "100%",
  fig.align = "center"
)
```

# Load Packages

```{r load-packages}
library(tidyverse)
library(here)
library(lme4)
library(ggplot2)
library(lmerTest)
library(performance)
library(broom.mixed)
library(kableExtra)
```

# Import Data

Import the imputed vaccine data generated in the previous step and briefly inspect the data structure.

```{r import-data}
vaccine <- read_csv(here("data/analytical", "vaccine_clean_imputed.csv"))
glimpse(vaccine)
summary(vaccine)
```

# Feature Engineering

Before fitting the model, apply feature engineering to the required variables:

-   `spending_per_student`: derived by Public spending on education per student by state in 2016 (in \$1,000s) is derived by dividing `statespending2016` by `schagepop2016`

-   `spending_scaled`: Because `spending_per_student` exhibited substantial variability across schools, the variable was standardized (mean = 0, SD = 1) for analysis

-   `combined_type`: Rare school types observed in only one state **('BOCES,' 'Charter,' 'Kindergarten,' 'Nonpublic')** were collapsed into an **'Other'** category to prevent confounding with state random intercepts. 'Public' and 'Private' types remained unchanged.

```{r spending-per-student}
# Calculate spending_per_student (spending / school aged pop.)
vaccine$spending_per_student <- vaccine$statespending2016 / vaccine$schagepop2016

# Compute mean for spending_per_student
mean_spending <- mean(vaccine$spending_per_student, na.rm = TRUE)

# distribution of Spending per Student
ggplot(vaccine,aes(x=spending_per_student))+
  geom_histogram(bins=30,fill="steelblue",color="white") +
  geom_vline(xintercept = mean_spending, color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = mean_spending, y = Inf, 
           label = paste("Mean =", round(mean_spending, 2)), 
           vjust = 2, hjust = -0.1, color = "red", size = 4) +
  labs(
    title="Distribution of Education Spending per Student",
    x="Spending (in $1,000s)",
    y="Number of Schools"
  )+
  theme_minimal()
```

```{r spending-scaled}
# 2. Standardize the spending_per_student to have a mean of 0 and SD of 1
vaccine$spending_scaled <- as.numeric(scale(vaccine$spending_per_student))

# Compute mean for spending_scaled
mean_scaled <- mean(vaccine$spending_scaled, na.rm = TRUE)

# distribution of spending_scaled
ggplot(vaccine,aes(x=spending_scaled))+
  geom_histogram(bins=30,fill="steelblue",color="white") +
  geom_vline(xintercept = mean_scaled, color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = mean_scaled, y = Inf, 
           label = paste("Mean =", round(mean_scaled, 2)), 
           vjust = 2, hjust = -0.1, color = "red", size = 4) +
  labs(
    title="Distribution of Scaled Education Spending per Student",
    x="Scaled Spending",
    y="Number of Schools"
  )+
  theme_minimal()

```

```{r combined-type}
# Check how many distinct states are in each type
vaccine %>% group_by(type) %>% 
  summarise(n_state = n_distinct(state),n_school = n()) %>%
  kable()

# Combine rare school types (BOCES, Charter, Kindergarten, Nonpublic) into "Other" 
# to prevent confounding with state random intercepts
rare_type <- c("BOCES", "Charter", "Kindergarten", "Nonpublic")
vaccine <- vaccine %>%
  mutate(combined_type = ifelse(type %in% rare_type, "Other", type))%>%
  # Set Public as the reference category
  mutate(combined_type = fct_relevel(combined_type, "Public"))
```

# Primary Analysis: MMR Vaccination Rate

Fit Random Intercepts Linear Mixed Model:

-   **Primary outcome:** Log-reflected, imputed school-level MMR vaccination rate.

-   **Predictor:** Standardized education spending per student

-   **Covariate:** School types

-   **Random Effect:** State

**Log-Reflected Transformation Formula:** $mmr_{new} = \log((max(mmr) + 1) - mmr)$

Since the distribution of MMR is left-skewed, a reflection method with a log transformation is applied to correct the skewness.

*Note: In the reflected data, we need to reverse the sign of the coefficients when interpreting the results, and higher values in the output correspond to lower values in the original scale.*

```{r apply-reflect-log-transformation-mmr}
# Apply reflect method to solve the left-skewed problems of mmr
max_mmr <- max(vaccine$mmr_imp, na.rm = TRUE)
vaccine$mmr_reflected <- (max_mmr + 1) - vaccine$mmr_imp
vaccine$mmr_reflected_log <- log(vaccine$mmr_reflected)

# Explore the distribution of MMR after transformation
ggplot(vaccine, aes(x = mmr_reflected_log)) +
  geom_histogram(fill = 'steelblue', bins = 30, color = 'white') +
  labs(
    title = "Distribution of Imputed MMR After Log-Reflected Transformation",
    subtitle = paste("n =", nrow(vaccine)),
    x = "Log-reflected, Imputed MMR",
    x = "Number of Schools"
  ) +
  theme_minimal()
```

```{r fit-null-mmr-model}
# Fit the null model to calculate the baseline ICC later
# and to justify that state-level clustering exists
model_mmr_null <- lmer(mmr_reflected_log ~ 1 + (1 | state), data = vaccine)
summary(model_mmr_null)
```

```{r fit-mmr-model}
# Fit the model with covariates
model_mmr <- lmer(mmr_reflected_log ~ spending_scaled + combined_type + (1 | state),
                  data = vaccine)
summary(model_mmr)
```

# Secondary Analysis: Overall Vaccination Rate

Fit the Random Intercepts Linear Mixed Model:

-   **Secondary Outcome:** Log-reflected, imputed school-level Overall vaccination rate.

-   **Predictor:** Standardized education spending per student

-   **Covariate:** School types

-   **Random Effect:** State

**Log-Reflected Transformation Formula:** $Overall_{new} = \log((max(Overall) + 1) - Overall)$

Since the distribution of Overall is left-skewed, a reflection method with a log transformation is applied to correct the skewness.

*Note: In the reflected data, we need to reverse the sign of the coefficients when interpreting the results, and higher values in the output correspond to lower values in the original scale.*

```{r apply-reflect-log-transformation-overall}
# Apply reflect method to solve the left-skewed problems of overall
max_overall <- max(vaccine$overall_imp, na.rm = TRUE)
vaccine$overall_reflected <- (max_overall + 1) - vaccine$overall_imp
vaccine$overall_reflected_log <- log(vaccine$overall_reflected)

# Explore the distribution of Overall after transformation
ggplot(vaccine, aes(x = overall_reflected_log)) +
  geom_histogram(fill = 'steelblue', bins = 30, color = 'white') +
  labs(
    title = "Distribution of Imputed Overall After Log-Reflected Transformation",
    subtitle = paste("n =", nrow(vaccine)),
    x = "Log-reflected, Imputed Overall",
    x = "Number of Schools"
  ) +
  theme_minimal()
```

```{r fit-null-overall-model}
# Fit the null model for calculate baseline of ICC 
# to justify that state-level clustering exists
model_overall_null <- lmer(overall_reflected_log ~ 1 + (1 | state), data = vaccine)
summary(model_overall_null)
```

```{r fit-overll-model}
# Fit the model with covariates
model_overall <- lmer(overall_reflected_log ~ spending_scaled + combined_type + 
                        (1 | state), data = vaccine)
summary(model_overall)
```

# Sensitivity Analysis: Stratified by School Type

Stratifying the models by school type to assess if the association differs across these groups

## Stratified MMR Models

### Public Schools Only (MMR)

```{r stratified-public-mmr}
model_mmr_public <- lmer(mmr_reflected_log ~ spending_scaled + (1 | state), 
                     data = filter(vaccine, combined_type == "Public"))
summary(model_mmr_public)
```

### Private Schools Only (MMR)

```{r stratified-private-mmr}
model_mmr_private <- lmer(mmr_reflected_log ~ spending_scaled + (1 | state), 
                      data = filter(vaccine, combined_type == "Private"))
summary(model_mmr_private)
```

### 'Other' Schools ("BOCES", "Charter", "Kindergarten", "Nonpublic") (MMR)

```{r stratified-other-mmr}
model_mmr_other <- lmer(mmr_reflected_log ~ spending_scaled + (1 | state), 
                           data = filter(vaccine, combined_type == "Other"))
summary(model_mmr_other)
```

## Stratified Overall Models

### Public Schools Only (Overall)

```{r stratified-public-overall}
model_overall_public <- lmer(overall_reflected_log ~ spending_scaled + (1 | state), 
                             data = filter(vaccine, combined_type == "Public"))
summary(model_overall_public)
```

### Private Schools Only (Overall)

```{r stratified-private-overall}
model_overall_private <- lmer(overall_reflected_log ~ spending_scaled + (1 | state), 
                              data = filter(vaccine, combined_type == "Private"))
summary(model_overall_private)
```

### 'Other' Schools ("BOCES", "Charter", "Kindergarten", "Nonpublic") (MMR)

```{r stratified-other-overall}
model_overall_other <- lmer(overall_reflected_log ~ spending_scaled + (1 | state), 
                              data = filter(vaccine, combined_type == "Other"))
summary(model_overall_other)
```

# Sensitivity Analysis: Stratified by Cohort

Stratifying the models by cohort (2017-2018 vs. 2018-2019) to assess if the association differs across cohorts

## Stratified MMR Models

### 2017-2018 Cohort Only (MMR)

```{r stratified-1718-mmr}
model_mmr_1718 <- lmer(mmr_reflected_log ~ spending_scaled + combined_type + (1 | state),
                   data = filter(vaccine, year == "2017" | year == "2017-18"))
summary(model_mmr_1718)
```

### 2018-2019 Cohort Only (MMR)

```{r stratified-1819-mmr}
model_mmr_1819 <- lmer(mmr_reflected_log ~ spending_scaled + combined_type + (1 | state),
                   data = filter(vaccine, year == "2018-19"))
summary(model_mmr_1819)
```

## Stratified Overall Models

### 2017-2018 Cohort Only (Overall)

```{r stratified-1718-overall}
model_overall_1718 <- lmer(overall_reflected_log ~ spending_scaled + combined_type + 
                             (1 | state), 
                           data = filter(vaccine, year == "2017" | year == "2017-18"))
summary(model_overall_1718)
```

### 2018-2019 Cohort Only (Overall)

```{r stratified-1819-overall}
model_overall_1819 <- lmer(overall_reflected_log ~ spending_scaled + combined_type + 
                             (1 | state), data = filter(vaccine, year == "2018-19"))
summary(model_overall_1819)
```

# Missing data analysis: Representativeness Check

Check the potential bias from missing data

```{r create-missing-outcome-indicator}
# Create indicator for missing outcome
vaccine <- vaccine %>%
  mutate(
    missing_mmr = is.na(mmr),
    missing_overall = is.na(overall))
```

## Missing MMR

### Compare spending_scaled between complete and missing MMR groups

```{r compare-complete-and-missing-mmr}
vaccine %>%
  group_by(missing_mmr) %>%
  summarise(
    n = n(),
    mean_spending = mean(spending_scaled, na.rm = TRUE),
    sd_spending = sd(spending_scaled, na.rm = TRUE),
    n_state = n_distinct(state),
    n_type = n_distinct(combined_type))
```

### t-test for spending difference

```{r t-test-mmr}
t.test(spending_scaled ~ missing_mmr, data = vaccine)
```

### Pearson's Chi-squared test

```{r chi-squared-test-mmr}
# School type distribution by missing MMR
table(vaccine$combined_type, vaccine$missing_mmr, useNA = "ifany")

# Pearson's Chi-squared test
chisq.test(table(vaccine$combined_type, vaccine$missing_mmr))
```

### Mixed-effects logistic regression

```{r mixed-effects-logistic-regression-mmr}
# Mixed-effects logistic regression to predict missing MMR
missing_mmr_glmer <- glmer(missing_mmr ~ spending_scaled + combined_type + 
                             (1|state),
                       data = vaccine, family = binomial)
summary(missing_mmr_glmer)
```

## Missing Overall

### Compare spending_scaled between complete and missing Overall groups

```{r compare-complete-and-missing-overall}
vaccine %>%
  group_by(missing_overall) %>%
  summarise(
    n = n(),
    mean_spending = mean(spending_scaled, na.rm = TRUE),
    sd_spending = sd(spending_scaled, na.rm = TRUE),
    n_state = n_distinct(state),
    n_type = n_distinct(combined_type))
```

### t-test for spending difference

```{r t-test-overall}
t.test(spending_scaled ~ missing_overall, data = vaccine)
```

### Pearson's Chi-squared test

```{r chi-squared-test-overall}
# Combined School type distribution by missing Overall
table(vaccine$combined_type, vaccine$missing_overall, useNA = "ifany")

# Pearson's Chi-squared test
chisq.test(table(vaccine$combined_type, vaccine$missing_overall))
```

### Mixed-effects logistic regression

```{r mixed-effects-logistic-regression-overall}
# Mixed-effects logistic regression to predict missing Overall
missing_overall_glmer <- glmer(missing_overall ~ spending_scaled + combined_type 
                               + (1|state),
                       data = vaccine, family = binomial)
summary(missing_overall_glmer)
```

# Model Validation

## Intraclass Correlation Coefficient (ICC)

Calculate Intraclass Correlation Coefficient (ICC) to check the percentage of total variation in the outcome attributable to the random intercept (state)

**Baseline ICC for MMR Model:**

```{r icc-mmr-null}
icc(model_mmr_null)
```

**ICC for MMR Model:**

```{r icc-mmr}
icc(model_mmr)
```

**Baseline ICC for Overall Model:**

```{r icc-overall-null}
icc(model_overall_null)
```

**ICC for Overall Model:**

```{r icc-overall}
icc(model_overall)
```

## Check residual normality and homoscedasticity

**Check residual normality and homoscedasticity for MMR model.**

```{r residual-mmr}
# plot fitted values vs. residuals for homoscedasticity
plot(model_mmr, which = 2)

# Q-Q plot for normality of residuals
qqnorm(resid(model_mmr), main = "Normal Q-Q Plot")
qqline(resid(model_mmr), col = "red")
```

**Check residual normality and homoscedasticity for Overall model.**

```{r residual-overall}
# plot fitted values vs. residuals for homoscedasticity
plot(model_overall, which = 2)

# Q-Q plot for normality of residuals
qqnorm(resid(model_overall), main = "Normal Q-Q Plot")
qqline(resid(model_overall), col = "red")
```

## Check for multicollinearity

**Check multicollinearity for MMR model.**

```{r check-multicollinearity-mmr}
check_collinearity(model_mmr)
```

**Check multicollinearity for Ovreall model.**

```{r check-multicollinearity-overall}
check_collinearity(model_overall)
```

# Create Table

## Define the helper functions

Define the helper functions to extract and format lmer model statistics

```{r helper-functions}
# 1. Helper to format Descriptive Stats
fmt_d <- function(x) sprintf("%.1f (%.1f) [%.1f, %.1f]", 
                             mean(x, na.rm=T), sd(x, na.rm=T), 
                             min(x, na.rm=T), max(x, na.rm=T))

# 2. Helper to extract model data and build the exact column structure
get_col <- function(mod, outcome_var) {
  t <- broom.mixed::tidy(mod, conf.int = TRUE)
  
  # Format p-values (<0.001 or 3 decimal places)
  p_vals <- ifelse(t$p.value[1:5] < 0.001, "< 0.001", sprintf("%.3f", t$p.value[1:5]))
  
  tibble(
    Variable = c(
      "Vaccination Rate", "Original Spending per Student (in $1,000s)", 
      "Intercept", "Education Spending Per Student (Scaled)",
      "School Type: Other", "School Type: Private", "School Type: Unknown",
      "Random Effect: State (SD)", "Residual (SD)", 
      "Observations (Schools)", "Number of States", "ICC (Adjusted)"
    ),
    Beta = c(
      fmt_d(outcome_var), fmt_d(vaccine$spending_per_student),                     
      sprintf("%.3f (%.3f)", t$estimate[1:5], t$std.error[1:5]),               
      sprintf("%.3f", t$estimate[6:7]),                                        
      nobs(mod), lme4::ngrps(mod)[["state"]], sprintf("%.3f", icc(mod)$ICC_adjusted)
    ),
    CI = c("", "", sprintf("[%.3f, %.3f]", t$conf.low[1:5], 
                           t$conf.high[1:5]), rep("", 5)),
    Pval = c("", "", p_vals, rep("", 5))
  )
}
```

## Apply helper to both models and merge them side-by-side

```{r apply-helper-to-models}
final_df <- left_join(
  get_col(model_mmr, vaccine$mmr_imp) %>% 
    rename(Beta_M = Beta, CI_M = CI, Pval_M = Pval),
  get_col(model_overall, vaccine$overall_imp) %>% 
    rename(Beta_O = Beta, CI_O = CI, Pval_O = Pval),
  by = "Variable"
)
```

## Generate table 2

```{r table-2}
kbl(final_df, 
    caption = paste("Table 2: Random Intercepts Linear Mixed Model Estimates for", 
                    "Log-Reflected MMR and Overall Vaccination Rates"),
    col.names = c("Variable", "Beta (SE)", "95% CI", "p-value", 
                  "Beta (SE)", "95% CI", "p-value"), 
    align = "lrrrrrr", booktabs = TRUE) %>%
  kable_classic(full_width = F, html_font = "Arial") %>%
  add_header_above(c(" " = 1, "MMR Vaccination Rate" = 3, 
                     "Overall Vaccination Rate" = 3)) %>%
  pack_rows("Descriptive Statistics: Mean (SD) [Min, Max]", 1, 2) %>%
  pack_rows("Fixed Effects", 3, 7) %>%
  pack_rows("Random Effects", 8, 9) %>%
  pack_rows("Model Fit & Validation", 10, 12) %>%
  column_spec(1, bold = T) %>%
  footnote(
    general = 
      c(paste("Note: Modeled outcomes were heavily left-skewed and transformed using", 
              "log((Max + 1) - Rate).", 
              "Consequently, a negative Beta indicates an INCREASE in the", 
              "original vaccination rate."),
               paste("Education Spending Per Student was standardized (mean = 0, SD = 1);", 
                     "the Beta represents the effect of a one standard", 
                     "deviation increase."),
                "The reference category for School Type is 'Public'."),
    general_title = ""
  ) %>%
  kable_styling(latex_options = c("scale_down", "Hold_position"))
```

\newpage

# Conclusion

Overall, the first step involved engineering the data by calculating and standardizing education spending per student, grouping rare school types into a single category ('Other'), and applying a log-reflected transformation to correct severe left-skewness in the MMR and overall vaccination outcomes. Next, random-intercept linear mixed models (LMMs) were fit and evaluated to assess the association between standardized education spending and vaccination rates, accounting for school type and clustering by state. Finally, two sensitivity analyses (stratified by school type and cohort) and missing data analyses (using t-tests, Pearson's Chi-squared tests, and mixed-effects logistic regression to check for missingness bias) were conducted. The models were then validated, and Table 2 was created to summarize the results in a report-ready format.
